Ship-based lidar evaluation of Southern Ocean clouds in the storm-resolving general circulation model ICON, and the ERA5 and MERRA-2 reanalyses

### 2024-12-13 manuscript.tex
Update on Overleaf.

Global storm-resolving models (GSRMs) are the next avenue of climate modelling.
Among them is a 5-km version of the Icosahedral Nonhydrostatic Weather and
Climate Model (ICON). The high resolution means that parameterizations of
convection and clouds including subgrid-scale clouds are forgone, relying on
explicit their simulation, but still using microphysics and turbulence
parametrization.  Standard-resolution (10--100-km) models, which use
subgrid-scale convection and cloud parametrization, have substantial cloud
biases over the Southern Ocean (SO), affecting radiation and sea surface
temperature.  We evaluated SO clouds in ICON and the ERA5 and MERRA-2
reanalyses.  The SO is dominated by low clouds, which cannot be observed
accurately from space due to overlapping clouds, attenuation, and ground
clutter.  Instead, we analysed about 2400 days of lidar observations from 31
voyages and a station during 2010--2021 using a ground-based lidar simulator.
ICON and the reanalyses underestimate the total cloud fraction by about 10 and
20\%, respectively. ICON and ERA5 overestimate the cloud occurrence peak at
about 500 m, potentially explained by their lifting condensation levels being
too high.  The reanalyses strongly underestimate near-surface clouds or fog.
MERRA-2 tends to underestimate cloud occurrence at all heights.  In daily cloud
cover, ICON and the reanalyses tend to be about 1 and 2 oktas clearer,
respectively.  Conditions of weak stability are associated with the largest
errors in ICON and the reanalyses.  Compared to radiosondes, potential
temperature is accurate in the reanalyses to about 0.5 K, but ICON
underestimates stability over the low-latitude SO and too humid in the boundary
layer. MERRA-2 is too humid at all heights. SO cloud biases remain a
substantial issue in the GSRM, but are an improvement over the lower-resolution
reanalyses. Explicitly resolved convection and cloud processes (except
microphysics) were not enough to address the model cloud biases.

@@

Increasing climate model resolution is one way of improving model accuracy of
representation of the climate system \citep{mauritsen2022}. It has been
practiced since the advent of climate modelling as more computational power,
memory, and storage capacity become available. It is, however, often not as
easy as changing the grid size because of the complex interplay between model
dynamics and physics, which necessitates adjusting and tuning all components
together.  Increasing resolution is of course limited by the available
computational power and a trade-off with increasing parameterization
complexity, which is another way of improving model accuracy.  Current
computational availability and acceleration from general-purpose computing on
graphics processing units (GPUs) is progressing to enable km-scale (also called
k-scale) Earth system models (ESMs) and coupled atmosphere--ocean general
circulation models (AOGCMs) in research conditions today and operationally in
the forthcoming years.  Therefore, it represents a natural advance in climate
modelling.  Global storm-resolving models (GSRMs) are emerging as a new front
in the development of high-resolution global climate models, with horizontal
grid resolutions of about 2--8 km \citep{satoh2019,stevens2019}. This is enough
to resolve mesoscale convective storms, but smaller-scale convective plumes and
cloud structure remain unresolved. At an approximately 5-km scale,
non-hydrostatic processes also become important \citep{weisman1997}, and for
this reason such models are generally non-hydrostatic. The terms global
cloud-resolving models or global convection-permitting/-resolving models are
also sometimes used interchangeably with GSRMs but imply that clouds or
convection are resolved explicitly, which is not entirely true for GSRMs, as
this would require an even higher horizontal resolution \citep{satoh2019}.
Representative of these efforts is the DYnamics of the Atmospheric general
circulation Modeled On Non-hydrostatic Domains (DYAMOND) project
\citep{stevens2019,dyamond}, which is an intercomparison of nine global GSRMs
over two 40-day time periods in summer (1 August -- 10 September 2016) and
winter (20 January -- 1 March 2020). A new one-year GSRM intercomparison is
currently proposed by \cite{takasuka2024}, with the hope of also evaluating the
seasonal cycle and large-scale circulation.  An alternative to using a
computationally costly GSRM is to train an artificial neural network on GSRM
output and use it for subgrid-scale clouds, as done with the GSRM ICON by
\cite{grundner2022} and \cite{grundner2023}.

nextGEMS is a project \citep{nextgems} focused on the research and development
of GSRMs at multiple modelling centres and universities in Europe.  The project
also develops GSRM versions of the Icosahedral Nonhydrostatic Weather and
Climate Model (ICON; \cite{hohenegger2023}), the Integrated Forecasting System
[IFS; \cite{ifs2023}], and their ocean components at eddy-resolving
resolutions: ICON-O \citep{korn2022} coupled with ICON and
Finite-Element/volumE Sea ice-Ocean Model [FESOM; \cite{wang2014}] and Nucleus
for European Modelling of the Ocean [NEMO; \cite{madec2023}] coupled with IFS.
The project has so far produced ICON and IFS simulations with three development
versions called Cycle 1--3 and a pre-final version, with a final production
version planned by the end of the project. nextGEMS is not the only project
developing GSRMs; other GSRMs (or GSRM versions of climate models) currently in
development include: Convection-Permitting Simulations With the E3SM Global
Atmosphere Model [SCREAM; \cite{caldwell2021}], Atmospheric Model [NICAM;
\cite{satoh2008}], Unified Model (UM), eXperimental System for High-resolution
modeling for Earth-to-Local Domain [X-SHiELD; \cite{shield}], Action de
Recherche Petite Echelle Grande Echelle-NonHydrostatic version [ARPEGE-NH;
\cite{bubnova1995,voldoire2017}], Finite-Volume Dynamical Core on the Cubed
Sphere [FV3, \cite{lin2004}], the National Aeronautics and Space Administration
(NASA) Goddard Earth Observing System global atmospheric model version 5
[GEOS5; \cite{putman2011}], Model for Prediction Across Scales [MPAS;
\cite{skamarock2012}], and System for Atmospheric Modeling [SAM;
\cite{khairoutdinov2003}].

@@

The Southern Ocean (SO) is known to be a problematic region for climate model
biases \citep{schuddeboom2021,hyder2018,cesana2022,zhao2022} due to a lack of
surface and in situ observations and being a lower priority region for
numerical weather prediction (NWP) and climate model development because of its
distance from populated areas.  Nevertheless, radiation biases and changes over
an area of its size have a substantial influence on the global climate
\citep{rintoul2011}, such as affecting the Earth radiation balance, ocean heat
and carbon uptake \citep{williams2023}, and the SO is an important part of the
global ocean conveyor belt \citep{wang2014}. In general, marine clouds have a
disproportionate effect on top of atmosphere (TOA) SW radiation due to the
relatively low albedo of the sea surface.  The relative longitudinal symmetry
of the SO means that model cloud biases tend to be similar across longitudes.

Here, we conventionally refer to the SO as ocean regions south of 40°S,
low-latitude SO as 40--55°S and high-latitude SO as south of 55°S. The reason
for this dividing latitude is to split the SO into about two equal zones, as
well as the results by \cite{schuddeboom2021} (Fig. 2b), showing a contrast in
CMIP model radiation biases, \cite{schuddeboom2019} (Fig. 2) and
\cite{kuma2020} (Fig. 3) showing contrasting radiation biases in the Hadley
Centre Global Environmental Model, and \cite{cesana2022} showing contrasting
cloud biases due to the 0°C isotherm reaching the surface at 55°S. The findings
of \cite{niu2024}, however, support a different dividing line of 62°S based on
cloud condensation nuclei concentration.

SO radiation biases have been relatively large and systematic compared to the
rest of the globe since at least CMIP3 \citep{trenberth2010}, and the SO SW
cloud radiative effect (CRE) bias is still positive in eight analysed CMIP6
models analysed by \cite{schuddeboom2021} over the high-latitude SO, whereas
over the low-latitude SO it tends to be more neutral or negative in some
models. Too much absorbed SW radiation over the SO was also identified in the
GSRM SCREAM \cite{caldwell2021}. Compensating biases are possible, such as the
`too few too bright' cloud bias, characterised by too small cloud fraction and
too large cloud albedo \citep{wall2017,kuma2020}, previously described by
\cite{webb2001}, \cite{weare2004}, \cite{zhang2005}, \cite{karlsson2008},
\cite{nam2012}, \cite{klein2013}, and \cite{bender2017} in other regions and
models, which means that a model can maintain a reasonable SW radiation balance
by reflecting too much SW radiation from clouds, but these cover too small
area.  A study by \cite{konsta2022} showed that this type of bias is still
present in six analysed CMIP6 models in tropical marine clouds, using the
GCM-Oriented Cloud-Aerosol Lidar and Infrared Pathfinder Satellite Observation
(CALIPSO) Cloud Product [CALIPSO--GOCCP; \cite{chepfer2010}] and Polarization
\& Anisotropy of Reflectances for Atmospheric Sciences coupled with
Observations from a Lidar [PARASOL; \cite{lier2008}] as a reference. They
suggest improper simulation of subgrid-scale cloud heterogeneity as a cause.
Compensating cloud biases in the Australian Community Climate and Earth System
Simulator (ACCESS) – Atmosphere-only model version 2 (AM2) over the SO were
analysed by \cite{fiddes2022} and \cite{fiddes2024}.  \cite{possner2022} showed
that over the SO, the DYAMOND GSRM ICON underestimates low-level cloud fraction
on the order of 30\% [relative to Moderate Resolution Imaging Spectroradiometer
(MODIS) data] and overestimates downwelling TOA SW radiation by approximately
10 Wm$^\mathrm{-2}$ [relative to the Clouds and the Earth’s Radiant Energy
System (CERES) data] in the highest model resolution run (2.5 km).
\cite{zhao2022} reported a similar SW radiation bias in five analysed CMIP6
models over high-latitude SO and the total cloud fraction underestimation on
the order of 10\% over the entire 40--60°S SO. Recently, \cite{ramadoss2024}
analysed 48 hours of km-scale ICON limited area model NWP simulations over a SO
region adjacent to Tasmania against the Clouds, Aerosols, Precipitation,
Radiation, and atmospherIc Composition Over the southeRn oceaN (CAPRICORN)
voyage cloud and precipitation observations \cite{mcfarquhar2021}. They found
the ICON cloud optical thickness was underestimated relative to Himawari‐8
satellite observations, but also identified large differences in cloud top
phase.

In general, sea surface temperature (SST) biases in the SO can originate either
in the atmosphere \citep{hyder2018}, caused by too much shortwave heating of
the surface, too little longwave cooling of the surface, or in the ocean
circulation.  Interactions of both are also possible, for example SST affecting
clouds and clouds affecting the surface radiation. Using the European Centre
for Medium-Range Weather Forecasts (ECMWF) Reanalysis 5 (ERA5) as a reference,
\cite{zhang2023} have shown that SST biases have improved in CMIP6 compared to
CMIP5, with SST overall increasing in the later CMIP phase. However, over the
SO this resulted in an even higher positive bias, especially in the Atlantic
Ocean (AO) sector of the SO, increasing by up to 1°C.  \cite{luo2023}
identified that the SO SST bias in an ensemble of 18 CMIP6 models originates
not from the surface heat and radiation fluxes (using reanalyses as a
reference), but from a warm bias in the Northern Atlantic Deep Water.

The main aim of this study is to evaluate the GSRM version of ICON. ICON
developed and maintained jointly by Deutscher Wetterdienst,
Max-Planck-Institute for Meteorology, Deutsches Klimarechenzentrum (DKRZ),
Karlsruhe Institute of Technology, and the Center for Climate Systems Modeling.
Previous studies have identified substantial large-scale biases in climate
model clouds over the SO, affecting sea surface temperature and the Earth’s
albedo. Our aim is to quantify how well the GSRM ICON is simulating clouds in
this region, particularly in light of the fact that subgrid-scale clouds and
convection are not parameterized in this model. This region is mostly dominated
by boundary layer clouds generated by shallow convection, and these are
problematic to observe by spaceborne lidars and radars, which are affected by
attenuation by overlapping and thick clouds \citep{mace2009,medeiros2010} and
ground clutter \citep{marchand2008}, respectively.  Specifically, the radar on
CloudSat and lidar on CALIPSO (neither of which are nowoperational) are
affected by the abovementioned issues, resulting in a strong underestimation of
cloud occurrence below 2 km relative to ground-based lidar observations
\citep{mcerlich2021}.  We hypothesise that this, in turn, can lead to
systematic biases in low clouds in climate models, which are frequently
evaluated against CloudSat--CALIPSO products. Reanalyses can also suffer from
cloud biases, as these are usually parametrised in their atmospheric component,
and also in regions where input observations are sparse.  This makes them a
problematic reference for clouds over the SO, and any biases relative to a
reanalysis should be interpreted with caution. Instead, we chose to use a large
set of ship-based observations conducted with ceilometers and lidars on board
of the RV \emph{Polarstern} and other voyages and stations as a reference for
the model evaluation.

@@

Together, we analysed data from 31 voyages of RV \emph{Polarstern}, the
resupply vessel (RSV) \emph{Aurora Australis}, RV \emph{Tangaroa}, RV
\emph{Nathaniel B.  Palmer}, Her (now His) Majesty's New Zealand Ship (HMNZS)
\emph{Wellington} and one sub-Antarctic station (Macquarie Island) in the SO
south of 40°S between 2010 and 2021. Fig. \ref{fig:map} shows a map of the
voyages, Table \ref{tab:voyages} list the voyages, campaigns, and stations, and
Table \ref{tab:voyage-references} lists references where available. Altogether,
the voyages and station dataset comprised 2421 days of data south of 40°S, but
the availability of ceilometer data was slightly shorter due to gaps in
measurements.

@@

The Vaisala CL51 and CT25K (photos in Fig. \ref{fig:map}d, e) are ceilometers
operating at a near-infrared wavelength of 910 nm and 905 nm, respectively.
The CL51 can also be configured to emulate the Vaisala CL31. The maximum range
is 15.4 km (CL51), 7.7 km (CL31 emulation mode with 5 m vertical resolution),
and 7.5 km (CT25K). The vertical resolution is 10 m (5 m configurable) in CL51
and 30 m in CT25K observations. The sampling (temporal) resolution is
configurable, and in our datasets is approximately 6 s for CL51 on AA15‐16, 16
s for CT25K on MARCUS and MICRE, 36 s for CL51 on RV \emph{Polarstern}, and
about 2.37 s for CL51 with CL31 emulation on TAN1502. The wavelengths of 905
and 910 nm are affected by water vapour absorption of about 20\% in the
mid-latitudes \citep{wiegner2015,wiegner2019}, with 910 nm affected more
strongly, but we do not expect this to be a significant issue as explained in
\cite{kuma2021}.  The instrument data files containing raw uncalibrated
backscatter were first converted to Network Common Data Form (NetCDF) with
cl2nc (\url{https://github.com/peterkuma/cl2nc}) and then processed with the
ALCF (Section \ref{sec:alcf}) to produce absolutely calibrated attenuated
volume backscattering coefficient (AVBC), cloud mask, cloud occurrence by
height, and the total cloud fraction. Because the CT25K uses a very similar
wavelength to CL51, equivalent calculations as for CL51 were done assuming a
wavelength of 910 nm. The Vaisala CL51 and CT25K instruments were used on most
of the voyages and stations analysed here. Fig.  \ref{fig:example}a shows an
example of AVBC derived from the CL51 instrument data.

\begin{figure}[b!]
\centering
\includegraphics[width=\textwidth]{img/example.png}
\caption{
An example of attenuated volume backscattering coefficient (AVBC) \textbf{(a)}
measured by CL51 during 24 hours on the PS81/3 voyage and \textbf{(b)} an
equivalent AVBC simulated with the ALCF from ERA5 data during the same time
period. The red line identifies the cloud mask determined by the ALCF.
}
\label{fig:example}
\end{figure}

@@

The Automatic Lidar and Ceilometer Framework (ALCF) is a ground-based lidar
simulator and a tool for processing observed lidar data, supporting various
instruments and models \citep{kuma2021}. It performs radiative transfer
calculations to derive equivalent lidar AVBC in an atmospheric model, which can
then be compared with observed AVBC. For this purpose, it takes the cloud
fraction, liquid and ice mass mixing ratio, temperature, and pressure model
fields as an input and is run offline (on the model output rather than inside
the model code). The lidar simulator in the ALCF is based on the instrument
simulator Cloud Feedback Model Intercomparison Project (CFMIP) Observation
Simulator Package (COSP) \citep{bodas-salcedo2011}.  After AVBC is calculated,
a cloud mask, cloud occurrence by height, and the total cloud fraction are
determined. The ALCF has been used by several research teams for model and
reanalysis evaluation
\citep{kuma2020,kremser2021,guyot2022,pei2023,whitehead2023,mcdonald2024}.

@@

A coupled (atmosphere--ocean) GSRM version of the ICON model is in development
at the nextGEMS project \citep{hohenegger2023}. ICON is an exceptionally
versatile model, allowing for simulations ranging from coarse-resolution ESM
simulations, GSRM simulations, limited area model simulations, to large eddy
simulations (LES), for both weather prediction and climate projections. ICON
uses the atmospheric component ICON-A \citep{giorgetta2018}, whose physics is
derived from ECHAM6 \citep{stevens2013}, and the ocean component ICON-O
\citep{korn2022}. Earlier runs of the GSRM ICON from DYAMOND were evaluated by
\cite{mauritsen2022}.

@@

ERA5 \citep{era5} is a reanalysis produced by the ECMWF.  It is based on a
numerical weather prediction model IFS version CY41R2.  It uses the
\cite{tiedtke1993} prognostic cloud scheme and \cite{forbes2014} for
mixed-phase clouds. The horizontal resolution is 0.25° in latitude and
longitude, which is about 28 km in the North--South direction and 14 km in the
East--West direction at 60°S.  Internally, the model uses 137 vertical levels.
Here, we use output at 1-hourly instantaneous time intervals, except for
radiation quantities, which are accumulations (from these we calculate daily
means).  Vertically resolved quantities are made available on 37 pressure
levels.

>>>>

Global storm-resolving models (GSRMs) are the next avenue of climate modelling.
Among them is a 5-km horizontal resolution version of the Icosahedral Nonhydrostatic Weather and
Climate Model (ICON). This high resolution means that parameterizations of
convection and clouds including subgrid-scale clouds are not required, though microphysics and turbulence
parametrizations are still needed.  Standard-resolution (10--100-km) models, which use
subgrid-scale convection and cloud parametrization, have substantial cloud
biases over the Southern Ocean (SO), adversely affecting radiation and sea surface
temperature.  We evaluated SO clouds in ICON and the ERA5 and MERRA-2
reanalyses.  The SO is dominated by low clouds, which cannot be observed
accurately from space due to overlapping clouds, attenuation, and ground
clutter.  Instead, we analysed about 2400 days of lidar observations from 31
voyages and a station during 2010--2021 using a ground-based lidar simulator.
ICON and the reanalyses underestimate the total cloud fraction by about 10 and
20\%, respectively. ICON and ERA5 overestimate the cloud occurrence peak at
about 500 m, potentially explained by their lifting condensation levels being
too high.  The reanalyses strongly underestimate near-surface clouds or fog.
MERRA-2 tends to underestimate cloud occurrence at all heights.  In daily cloud
cover, ICON and the reanalyses tend to be about 1 and 2 oktas clearer,
respectively. The largest
errors in ICON and the reanalyses occur during periods of weak stability.  Compared to radiosondes, potential
temperature is accurate in the reanalyses to about 0.5 K, but ICON
underestimates stability over the low-latitude SO and is too humid in the boundary
layer. MERRA-2 is too humid at all heights. SO cloud biases remain a
substantial issue in the GSRM, but are an improvement over the lower-resolution
reanalyses. Explicitly resolved convection and cloud processes (except
microphysics) were not enough to address the model cloud biases.

@@

Increasing climate model resolution is one way of improving the accuracy of
the representation of the climate system in models \citep{mauritsen2022}. It has been
practiced since the advent of climate modelling as more computational power,
memory, and storage capacity become available. It is, however, often not as
easy as changing the grid size because of the complex interplay between model
dynamics and physics, which necessitates adjusting and tuning all components
together. Increasing resolution is of course limited by the available
computational power and a trade-off with increasing parameterization
complexity, which is another way of improving model accuracy. Current
computational availability and acceleration from general-purpose computing on
graphics processing units (GPUs) has progressed to enable km-scale (also called
k-scale) Earth system models (ESMs) and coupled atmosphere--ocean general
circulation models (AOGCMs) for research today and will become operational in
the future. Therefore, it represents a natural advance in climate
modelling. Global storm-resolving models (GSRMs) are emerging as a new front
in the development of high-resolution global climate models, with horizontal
grid resolutions of about 2--8 km \citep{satoh2019,stevens2019}. This resolution is enough
to resolve mesoscale convective storms, but smaller-scale convective plumes and
cloud structure remain unresolved. At an approximately 5-km scale,
non-hydrostatic processes also become important \citep{weisman1997}, and for
this reason such models are generally non-hydrostatic. The terms global
cloud-resolving models or global convection-permitting/-resolving models are
also sometimes used interchangeably with GSRMs but imply that clouds or
convection are resolved explicitly, which is not entirely true for GSRMs, as
this would require an even higher horizontal resolution \citep{satoh2019}.
Representative of these efforts is the DYnamics of the Atmospheric general
circulation Modeled On Non-hydrostatic Domains (DYAMOND) project
\citep{stevens2019,dyamond}, which is an intercomparison of nine global GSRMs
over two 40-day time periods in summer (1 August -- 10 September 2016) and
winter (20 January -- 1 March 2020). A new one-year GSRM intercomparison is
currently proposed by \cite{takasuka2024}, with the hope of also evaluating the
seasonal cycle and large-scale circulation.  An alternative to using a
computationally costly GSRM is to train an artificial neural network on GSRM
output and use it for subgrid-scale clouds, as done with the GSRM ICON by
\cite{grundner2022} and \cite{grundner2023}.

The nextGEMS project \citep{nextgems} focuses on the research and development
of GSRMs at multiple modelling centres and universities in Europe. The project
also develops GSRM versions of the Icosahedral Nonhydrostatic Weather and
Climate Model (ICON; \cite{hohenegger2023}), the Integrated Forecasting System
[IFS; \cite{ifs2023}], and their ocean components at eddy-resolving
resolutions: ICON-O \citep{korn2022} coupled with ICON and
Finite-Element/volumE Sea ice-Ocean Model [FESOM; \cite{wang2014}] and Nucleus
for European Modelling of the Ocean [NEMO; \cite{madec2023}] coupled with IFS.
The project has so far produced ICON and IFS simulations with three development
versions called Cycle 1--3 and a pre-final version, with a final production
version planned by the end of the project. nextGEMS is not the only project
developing GSRMs; other GSRMs (or GSRM versions of climate models) currently in
development include: Convection-Permitting Simulations With the E3SM Global
Atmosphere Model [SCREAM; \cite{caldwell2021}], Atmospheric Model [NICAM;
\cite{satoh2008}], Unified Model (UM), eXperimental System for High-resolution
modeling for Earth-to-Local Domain [X-SHiELD; \cite{shield}], Action de
Recherche Petite Echelle Grande Echelle-NonHydrostatic version [ARPEGE-NH;
\cite{bubnova1995,voldoire2017}], Finite-Volume Dynamical Core on the Cubed
Sphere [FV3, \cite{lin2004}], the National Aeronautics and Space Administration
(NASA) Goddard Earth Observing System global atmospheric model version 5
[GEOS5; \cite{putman2011}], Model for Prediction Across Scales [MPAS;
\cite{skamarock2012}], and System for Atmospheric Modeling [SAM;
\cite{khairoutdinov2003}].

@@

The Southern Ocean (SO) is known to be a problematic region for climate model
biases \citep{schuddeboom2021,hyder2018,cesana2022,zhao2022} due to a lack of
surface and in situ observations and being a lower priority region for
numerical weather prediction (NWP) and climate model development because of its
distance from populated areas. Nevertheless, radiation biases and changes over
an area of its size have a substantial influence on the global climate
\citep{rintoul2011}, such as affecting the Earth radiation balance, ocean heat
and carbon uptake \citep{williams2023}, and the SO is also an important part of the
global ocean conveyor belt \citep{wang2014}. In general, marine clouds have a
disproportionate effect on top of atmosphere (TOA) SW radiation due to the
relatively low albedo of the sea surface.  The relative longitudinal symmetry
of the SO means that model cloud biases tend to be similar across longitudes.

Here, we refer to the SO as ocean regions south of 40°S,
low-latitude SO as 40--55°S and high-latitude SO as south of 55°S. The reason
for this dividing latitude is to split the SO into about two equal zones, as
well as the results by \cite{schuddeboom2021} (Fig. 2b) which show a contrast in
CMIP model radiation biases. \cite{schuddeboom2019} (Fig. 2) and
\cite{kuma2020} (Fig. 3) also show contrasting radiation biases in the Hadley
Centre Global Environmental Model, which is also supported by \cite{cesana2022} which  displays contrasting cloud biases due to the 0°C isotherm reaching the surface at 55°S. The findings
of \cite{niu2024}, however, support a different dividing line of 62°S based on
cloud condensation nuclei concentration.

SO radiation biases have been relatively large and systematic compared to the
rest of the globe since at least CMIP3 \citep{trenberth2010}, and the SO SW
cloud radiative effect (CRE) bias is still positive in eight analysed CMIP6
models analysed by \cite{schuddeboom2021} over the high-latitude SO, whereas
over the low-latitude SO it tends to be more neutral or negative in some
models. Too much absorbed SW radiation over the SO was also identified in the
GSRM SCREAM \cite{caldwell2021}. Compensating biases are possible, such as the
`too few too bright' cloud bias, characterised by too small cloud fraction and
too large cloud albedo \citep{wall2017,kuma2020}, previously described by
\cite{webb2001}, \cite{weare2004}, \cite{zhang2005}, \cite{karlsson2008},
\cite{nam2012}, \cite{klein2013}, and \cite{bender2017} in other regions and
models, which means that a model can maintain a reasonable SW radiation balance
by reflecting too much SW radiation from clouds, but these cover too small
area. A study by \cite{konsta2022} showed that this type of bias is still
present in six analysed CMIP6 models in tropical marine clouds, using the
GCM-Oriented Cloud-Aerosol Lidar and Infrared Pathfinder Satellite Observation
(CALIPSO) Cloud Product [CALIPSO--GOCCP; \cite{chepfer2010}] and Polarization
\& Anisotropy of Reflectances for Atmospheric Sciences coupled with
Observations from a Lidar [PARASOL; \cite{lier2008}] as a reference. They
suggest improper simulation of subgrid-scale cloud heterogeneity as a cause.
Compensating cloud biases in the Australian Community Climate and Earth System
Simulator (ACCESS) – Atmosphere-only model version 2 (AM2) over the SO were
analysed by \cite{fiddes2022} and \cite{fiddes2024}.  \cite{possner2022} showed
that over the SO, the DYAMOND GSRM ICON underestimates low-level cloud fraction
on the order of 30\% [relative to Moderate Resolution Imaging Spectroradiometer
(MODIS) data] and overestimates downwelling TOA SW radiation by approximately
10 Wm$^\mathrm{-2}$ [relative to the Clouds and the Earth’s Radiant Energy
System (CERES) data] in the highest model resolution run (2.5 km).
\cite{zhao2022} reported a similar SW radiation bias in five analysed CMIP6
models over the high-latitude SO and an underestimation of total cloud fraction on
the order of 10\% over the entire 40--60°S SO. Recently, \cite{ramadoss2024}
analysed 48 hours of km-scale ICON limited area model NWP simulations over a SO
region adjacent to Tasmania against the Clouds, Aerosols, Precipitation,
Radiation, and atmospherIc Composition Over the southeRn oceaN (CAPRICORN)
voyage cloud and precipitation observations \cite{mcfarquhar2021}. They found
the ICON cloud optical thickness was underestimated relative to Himawari‐8
satellite observations, but also identified large differences in cloud top
phase.

In general, sea surface temperature (SST) biases in the SO can originate either
in the atmosphere \citep{hyder2018}, caused by too much shortwave heating of
the surface, too little longwave cooling of the surface, or in the ocean
circulation. Interactions of both are also possible, for example SST affecting
clouds and clouds affecting the surface radiation. Using the European Centre
for Medium-Range Weather Forecasts (ECMWF) Reanalysis 5 (ERA5) as a reference,
\cite{zhang2023} have shown that SST biases have improved in CMIP6 compared to
CMIP5, with SST overall increasing in CMIP6. However, over the
SO this resulted in an even higher positive bias, especially in the Atlantic
Ocean (AO) sector of the SO, increasing by up to 1°C.  \cite{luo2023}
identified that the SO SST bias in an ensemble of 18 CMIP6 models originates
not from the surface heat and radiation fluxes (using reanalyses as a
reference), but from a warm bias in the Northern Atlantic Deep Water.

The main aim of this study is to evaluate the GSRM version of ICON. ICON
is developed and maintained jointly by Deutscher Wetterdienst,
Max-Planck-Institute for Meteorology, Deutsches Klimarechenzentrum (DKRZ),
Karlsruhe Institute of Technology, and the Center for Climate Systems Modeling.
Previous studies have identified substantial large-scale biases in climate
model clouds over the SO, affecting sea surface temperature and the Earth’s
albedo. Our aim is to quantify how well the GSRM ICON simulates clouds in
this region, particularly in light of the fact that subgrid-scale clouds and
convection are not parameterized in this model. This region is mostly dominated
by boundary layer clouds generated by shallow convection, and these are
problematic to observe by spaceborne lidars and radars, which are affected by
attenuation by overlapping and thick clouds \citep{mace2009,medeiros2010} and
ground clutter \citep{marchand2008}, respectively.  Specifically, the radar on
CloudSat and lidar on CALIPSO (neither of which are now operational) are
affected by the above mentioned issues, resulting in a strong underestimation of
cloud occurrence below 2 km relative to ground-based lidar observations
\citep{mcerlich2021}.  We hypothesise that this, in turn, can lead to
systematic biases in low clouds in climate models, which are frequently
evaluated against CloudSat--CALIPSO products. Reanalyses can also suffer from
cloud biases, as these are usually parametrised in their atmospheric component,
and also in regions where input observations are sparse. This makes them a
problematic reference for clouds over the SO, and any biases relative to a
reanalysis should be interpreted with caution. Instead, we chose to use a large
set of ship-based observations conducted with ceilometers and lidars on board
the RV \emph{Polarstern} and other voyages and stations as a reference for model evaluation.

@@

Together, we analysed data from 31 voyages of RV \emph{Polarstern}, the
resupply vessel (RSV) \emph{Aurora Australis}, RV \emph{Tangaroa}, RV
\emph{Nathaniel B.  Palmer}, Her (now His) Majesty's New Zealand Ship (HMNZS)
\emph{Wellington} and one sub-Antarctic station (Macquarie Island) in the SO
south of 40°S between 2010 and 2021. Fig. \ref{fig:map} shows a map of the
voyages, Table \ref{tab:voyages} lists the voyages, campaigns, and stations, and
Table \ref{tab:voyage-references} lists references where available. Altogether,
the voyages and station dataset comprised 2421 days of data south of 40°S, but
the availability of ceilometer data was slightly shorter due to gaps in
measurements.

@@

The Vaisala CL51 and CT25K (photos in Fig. \ref{fig:map}d, e) are ceilometers
operating at a near-infrared wavelength of 910 nm and 905 nm, respectively.
The CL51 can also be configured to emulate the Vaisala CL31. The maximum range
is 15.4 km (CL51), 7.7 km (CL31 emulation mode with 5 m vertical resolution),
and 7.5 km (CT25K). The vertical resolution is 10 m (5 m configurable) in CL51
and 30 m in CT25K observations. The sampling (temporal) resolution is
configurable, and in our datasets is approximately 6 s for CL51 on AA15‐16, 16
s for CT25K on MARCUS and MICRE, 36 s for CL51 on RV \emph{Polarstern}, and
about 2.37 s for CL51 with CL31 emulation on TAN1502. The wavelengths of 905
and 910 nm are both affected by water vapour absorption of about 20\% in the
mid-latitudes \citep{wiegner2015,wiegner2019}, with 910 nm affected more
strongly, but we do not expect this to be a significant issue as explained in
\cite{kuma2021}.  The instrument data files containing raw uncalibrated
backscatter were first converted to Network Common Data Form (NetCDF) with
cl2nc (\url{https://github.com/peterkuma/cl2nc}) and then processed with the
ALCF (Section \ref{sec:alcf}) to produce absolutely calibrated attenuated
volume backscattering coefficient (AVBC), cloud mask, cloud occurrence by
height, and the total cloud fraction. Because the CT25K uses a very similar
wavelength to CL51, equivalent calculations as for CL51 were done assuming a
wavelength of 910 nm. The Vaisala CL51 and CT25K instruments were used on most
of the voyages and stations analysed here. Fig.  \ref{fig:example}a shows an
example of AVBC derived from the CL51 instrument data.

\begin{figure}[b!]
\centering
\includegraphics[width=\textwidth]{img/example.png}
\caption{
An example of attenuated volume backscattering coefficient (AVBC) \textbf{(a)}
measured by a CL51 during 24 hours on the PS81/3 voyage and \textbf{(b)} an
equivalent AVBC simulated with the ALCF from ERA5 data during the same time
period. The red line identifies the cloud mask determined by the ALCF.
}
\label{fig:example}
\end{figure}

@@

The Automatic Lidar and Ceilometer Framework (ALCF) is a ground-based lidar
simulator and a tool for processing observed lidar data, supporting various
instruments and models \citep{kuma2021}. It performs radiative transfer
calculations to derive equivalent lidar AVBC from an atmospheric model, which can
then be compared with observed AVBC. For this purpose, it takes the cloud
fraction, liquid and ice mass mixing ratio, temperature, and pressure model
fields as an input and is run offline (on the model output rather than inside
the model code). The lidar simulator in the ALCF is based on the instrument
simulator Cloud Feedback Model Intercomparison Project (CFMIP) Observation
Simulator Package (COSP) \citep{bodas-salcedo2011}.  After AVBC is calculated,
a cloud mask, cloud occurrence by height, and the total cloud fraction are
determined. The ALCF has been used by several research teams for model and
reanalysis evaluation
\citep{kuma2020,kremser2021,guyot2022,pei2023,whitehead2023,mcdonald2024}.

@@

A coupled (atmosphere--ocean) GSRM version of the ICON model is in development
as part of the nextGEMS project \citep{hohenegger2023}. ICON is an exceptionally
versatile model, allowing for simulations ranging from coarse-resolution ESM
simulations, GSRM simulations, limited area model simulations, to large eddy
simulations (LES), for both weather prediction and climate projections. ICON
uses the atmospheric component ICON-A \citep{giorgetta2018}, whose physics is
derived from ECHAM6 \citep{stevens2013}, and the ocean component ICON-O
\citep{korn2022}. Earlier runs of the GSRM ICON from DYAMOND were evaluated by
\cite{mauritsen2022}.

@@

ERA5 \citep{era5} is a reanalysis produced by ECMWF.  It is based on a
numerical weather prediction model IFS version CY41R2. It uses the
\cite{tiedtke1993} prognostic cloud scheme and \cite{forbes2014} for
mixed-phase clouds. The horizontal resolution is 0.25° in latitude and
longitude, which is about 28 km in the North--South direction and 14 km in the
East--West direction at 60°S.  Internally, the model uses 137 vertical levels.
Here, we use output at 1-hourly instantaneous time intervals, except for
radiation quantities, which are accumulations (from these we calculate daily
means).  Vertically resolved quantities are made available on 37 pressure
levels.
